---
title: "Using On-Body Sensors to Classify Weightlifting Errors"
author: "Jim Pfleger"
date: "January 24, 2016"
output: html_document
---

```{r load_libraries,cache=TRUE,echo=FALSE,results='hide',message=FALSE}
library(dplyr)
library(caret)
library(randomForest)
```

```{r load_data,dependson='load_libraries',cache=TRUE,echo=FALSE,results='hide'}
testing_raw <- read.csv('data/pml-testing.csv', stringsAsFactors = FALSE, na.strings = c("#DIV/0!", '""'))
testing_raw <- testing_raw %>% mutate(is_train = FALSE, classe = NA)

training_raw <- read.csv('data/pml-training.csv', stringsAsFactors = FALSE, na.strings = c("#DIV/0!", '""'))
training_raw <- training_raw %>% mutate(is_train = TRUE, problem_id = NA)
```


# Summary

When lifting weights, proper technique is important to both prevent injury and maximize the effectiveness of training. This paper presents a model for identifying proper and improper technique based on data from on-body sensors.

```{r clean_data,dependson='load_data',cache=TRUE,echo=FALSE,results='hide'}
### Merge data sets for clean up
merged <- rbind(testing_raw, training_raw)

merged <- merged %>%
  select(
    -starts_with('min'),
    -starts_with('max'),
    -starts_with('avg'),
    -starts_with('stddev'),
    -starts_with('var'),
    -starts_with('amplitude'),
    -starts_with('skewness'),
    -starts_with('kurtosis'),
    -one_of('X', 'user_name'),
    -matches('timestamp'),
    -matches('window')
  ) %>%
  mutate(
    classe = factor(classe)
  ) %>%
  mutate_each(funs(as.numeric), matches("arm")) %>%
  mutate_each(funs(as.numeric), matches("belt")) %>%
  mutate_each(funs(as.numeric), matches("dumbbell"))

training <- merged %>%
  filter(is_train == TRUE) %>%
  select(-is_train, -problem_id)

testing <- merged %>%
  filter(is_train == FALSE) %>%
  select(-is_train, -classe)

rm(training_raw, testing_raw, merged)
```

# Data Preparation

The model was trained on a subset of the Weight Lifting Exercises Dataset originally by Velloso _et al._ [^data] The classe column -- which identifies whether the exercise was performed properly, or which one of four errors occurred -- was kept and converted to a factor. The only other columns that were kept were those related to the raw readings from the arm, belt, and dumbbell sensors.

[^data]: Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. [Qualitative Activity Recognition of Weight Lifting Exercises](http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201). Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13). Stuttgart, Germany: ACM SIGCHI, 2013.

```{r create_model,dependson='clean_data',cache=TRUE,cache.lazy=FALSE,echo=FALSE}
set.seed(130991)

iter <- 10
ntree <- 25

ctrl <- trainControl(
  number = iter
)

predictors = select(training, -classe)
classe = training$classe

model <- train(
  x = predictors,
  y = classe,
  method = 'rf',
  preProcess = NULL,
  importance = TRUE,
  do.trace = FALSE,
  proximity = FALSE,
  ntree = ntree,
  trControl = ctrl)
```

# Model Selection

A random forest model was used because random forests have been shown to have superior performance in classification problems such as this. The model was trained on the entire training set (_n_ = `r nrow(training)`) using bootstraping to do cross-validation during training. `r iter` iterations of `r ntree` trees were run. Each iteration/tree was run with $m \in (`r model$results$mtry`)$ randomly selected predictors. The final model uses `r model$finalModel$mtry` predictors whose importance is shown below.

```{r plot_importance,dependson='create_model',cache=TRUE,echo=FALSE}
varImpPlot(model$finalModel, main="Variable Importance")
```

# Model Performance

# Results



create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did
